---
title: "Lifestyle and Well-being"
author: "Shanti Agung"
date: "4/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This mini project is an attempt to interpret machine learning models on well-being, mostly with the `iml` package. Well-being measures that I focus on are: stress level and the experience of flow. In particular, models are trained to predict whether a person falls within high-stress or low-stress category, and to predict hours of flow that a person experience in a week. 

## R packages

I use the following packages:

```{r}

# load required packages
library(tidyverse)   # data import, transformation, and visualization
library(lubridate)   # date wrangling
library(gbm)         # machine learning modeling
library(mlr)         # machine learning modeling
library(caret)       # machine learning modeling
library(iml)         # machine learning interpretation
library(lime)        # machine learning interpretation

```

## Prepare dataset

```{r}

# upload dataset
wellbeing <- read_csv("./data/Wellbeing_and_lifestyle_data_Kaggle.csv")

# remove observation with missing values
wellbeing <- na.omit(wellbeing) 

# tidy variables
wellbeing <- wellbeing %>% 
  mutate(DATE = str_extract(Timestamp, pattern = "^\\d+/\\d+/\\d+")) %>% 
  mutate(DATE = mdy(DATE))

age_levels <- c("Less than 20", "21 to 35", "36 to 50", "51 or more")

wellbeing <- wellbeing %>% 
  mutate(AGE = factor(AGE, levels = age_levels)) %>% 
  mutate(GENDER = factor(GENDER, levels = unique(GENDER)))

# create new variables
wellbeing <- wellbeing %>% 
  mutate(HI_STRESS = if_else(DAILY_STRESS > 3, 1, 0)) %>% 
  mutate(HI_STRESS = factor(HI_STRESS))

wellbeing <- wellbeing %>%
  mutate(YEAR = year(DATE)) %>%
  mutate(MONTH = month(DATE,label = TRUE)) %>%
  mutate(WEEK = wday(DATE, label = TRUE))

```

Because covid-19 pandemic affects our lifestyle and well-being in unusual ways, I use data prior to 2020.

```{r}

wellbeing <- wellbeing %>% 
  filter(YEAR < 2020)

dim(wellbeing)

glimpse(wellbeing)
```

Preparation of training and test sets for predicting high-stress: 

```{r}

# subset data for high-stress
wellbeing_01 <- wellbeing %>% 
  select(-c(Timestamp, DAILY_STRESS, FLOW, DAILY_SHOUTING, WORK_LIFE_BALANCE_SCORE, DATE))

# prepare train and test set for hi_stress
set.seed(2)
train_stress <- sample(1:nrow(wellbeing_01), nrow(wellbeing_01)/2)
wellbeing_01_test <- wellbeing_01[-train_stress,]
high_stress_test <- wellbeing_01$HI_STRESS[-train_stress]

```

Preparation of training and test sets for predicting flow experience: 

```{r}

# subset data for flow experience:
wellbeing_02 <- wellbeing %>% 
  select(-c(Timestamp, DAILY_STRESS, HI_STRESS, DAILY_SHOUTING, WORK_LIFE_BALANCE_SCORE, DATE))

# prepare train and test set for flow
set.seed(2)
train_flow <- sample(1:nrow(wellbeing_02), nrow(wellbeing_02)/2)
wellbeing_02_test <- wellbeing_02[-train_flow,]
flow_test <- wellbeing_02$FLOW[-train_flow]
```


## Train models using mlr package

Train a model to predict whether a person falls into high-stress category:
```{r}
set.seed(5)
hstress_task <- makeClassifTask(wellbeing_01[train_stress,],
                                     target = "HI_STRESS", id = 'hstress')
mod_hstress <- train(mlr::makeLearner(cl = 'classif.randomForest', id = 'hstress-rf',
                                           predict.type = 'prob'),
                          hstress_task)

pred_hstress <- Predictor$new(mod_hstress, data = wellbeing_01[train_stress,],
                                   class = "1")
```

Train a model to predict hours of flow that a person experience:
```{r}
set.seed(42)
flow_task <- makeRegrTask(data = wellbeing_02[train_flow,], target = "FLOW")
mod_flow <- train(makeLearner(cl = 'regr.randomForest', id = 'flow-rf'), flow_task)
pred_flow <- Predictor$new(mod_flow, wellbeing_02[train_flow,])
```

## Partial dependence plot (PDP)

PDP for high-stress prediction model and features of weekly mediation and lost vacation.

```{r}

hstress_pdp_1 <- iml::FeatureEffect$new(pred_hstress, "WEEKLY_MEDITATION", method = "pdp")$plot() +
  scale_y_continuous("Predicted high-stress probability") +
  scale_x_continuous("Weekly meditation (hours)")

hstress_pdp_2 <- iml::FeatureEffect$new(pred_hstress, "LOST_VACATION", method = "pdp")$plot() +
  scale_y_continuous("Predicted high-stress probability") +
  scale_x_continuous("Lost vacation per year (days)")

# plot pdp
hstress_pdp_1

hstress_pdp_2

```

PDP for flow prediction model and features of time for passion and achievement.

```{r}

flow_pdp_iml_1 <- FeatureEffect$new(pred_flow, feature = "TIME_FOR_PASSION",
                                         method = "pdp")$plot() +
  scale_x_continuous("Time for passion (hours)") +
  scale_y_continuous("Predicted daily flow (hours)")

flow_pdp_iml_2 <- FeatureEffect$new(pred_flow, feature = "ACHIEVEMENT",
                                         method = "pdp")$plot() +
  scale_x_continuous("Number of achievements (last 12 months)") +
  scale_y_continuous("Predicted daily flow (hours)")

# plot

flow_pdp_iml_1 

flow_pdp_iml_2

```


## Individual Conditional Expectation (ICE)

ICE for high-stress prediction model and features of weekly mediation and lost vacation.

```{r}

hstress_ice_1 <- FeatureEffect$new(pred_hstress, "WEEKLY_MEDITATION", method = "ice")$plot()
hstress_ice_2 <- FeatureEffect$new(pred_hstress, "LOST_VACATION", method = "ice")$plot()

# plot
hstress_ice_1
hstress_ice_2

```

Because the large numbers of observations, the ICE plots above are difficult to discern. Centered ICE plots of the model are:

```{r}
hstress_ice_1_centered <- FeatureEffect$new(pred_hstress, "WEEKLY_MEDITATION",
                                                 center.at = min(wellbeing_01$WEEKLY_MEDITATION[train_stress]),
                                                 method = "pdp+ice")$plot()

hstress_ice_2_centered <- FeatureEffect$new(pred_hstress, "LOST_VACATION",
                                                 center.at = min(wellbeing_01$LOST_VACATION[train_stress]),
                                                 method = "pdp+ice")$plot()

# plot
hstress_ice_1_centered +
  scale_y_continuous("High-stress prob. difference to 0 hrs weekly meditation")

hstress_ice_2_centered +
  scale_y_continuous("High-stress prob. difference to no lost vacation")

```

ICE for flow prediction model and features of time for passion and achievement.

```{r}

p1 <- FeatureEffect$new(pred_flow, "TIME_FOR_PASSION", method = "ice")$plot() +
  scale_x_continuous("Time for passion (hour)") +
  scale_y_continuous("Predicted flow experience (hours)")

p2 <- FeatureEffect$new(pred_flow, "ACHIEVEMENT", method = "ice")$plot() +
  scale_x_continuous("Number of achievements (last 12 months)") +
  scale_y_continuous("Predicted flow experience (hours)")

# plot
p1
p2

```

Centered ICE for flow prediction model:
```{r}

flow_ice_centered_1_iml <- FeatureEffect$new(pred_flow, feature = "TIME_FOR_PASSION",
                                                  center.at = min(wellbeing_02$TIME_FOR_PASSION[train_flow]),
                                                  method = "pdp+ice")$plot()

flow_ice_centered_2_iml <- FeatureEffect$new(pred_flow, feature = "ACHIEVEMENT",
                                                  center.at = min(wellbeing_02$ACHIEVEMENT[train_flow]),
                                                  method = "pdp+ice")$plot() 

# plot
flow_ice_centered_1_iml
flow_ice_centered_2_iml
```


## Accumulated Local Effects (ALE)

ALE for high-stress prediction model and features of weekly mediation and lost vacation.

```{r}

hstress_ale_1 <- FeatureEffect$new(pred_hstress, "WEEKLY_MEDITATION", method = "ale")$plot() +
  scale_y_continuous("ALE of high-stress") +
  scale_x_continuous("Weekly meditation (hours)")

hstress_ale_2 <- FeatureEffect$new(pred_hstress, "LOST_VACATION", method = "ale")$plot() +
  scale_y_continuous("ALE of high-stress") +
  scale_x_continuous("Lost vacation per year (days)")

# plot
hstress_ale_1
hstress_ale_2

```

ALE for flow prediction model and features of time for passion and achievement.

```{r}

flow_ale_1 <- FeatureEffect$new(pred_flow, feature = "TIME_FOR_PASSION",
                                     method = "ale")$plot() +
  scale_x_continuous("Time for passion (hours)") +
  scale_y_continuous("ALE of flow experience")

flow_ale_2 <- FeatureEffect$new(pred_flow, feature = "ACHIEVEMENT",
                                     method = "ale")$plot() +
  scale_x_continuous("Number of achievements (last 12 months)") +
  scale_y_continuous("ALE of flow experience")

# plot
flow_ale_1
flow_ale_1

```

## Permutation Feature Importance

Feature importance for high-stress prediction model.

```{r}
imp_hstress_ce <- FeatureImp$new(pred_hstress, loss = "ce")
plot(imp_hstress_ce)
```

Feature importance for flow prediction model. 

```{r}
imp_flow_mae <- FeatureImp$new(pred_flow, loss ="mae")
plot(imp_flow_mae)

```

## Local Surrogate (LIME)

At the moment, I have not been able applying LIME using the `iml` package. Therefore temporarily, I resort to `lime` package. So far, I've only managed to implement LIME to model flow experience. 

Below a flow model is trained using boosting from `gbm` package. The reference that I used to interpret a gbm model using the `lime` package is [here](http://uc-r.github.io/gbm_regression).

```{r}

# train model
set.seed(1)
boost_flow <- gbm(FLOW~., data = wellbeing_02[train_flow,], distribution = "poisson",
                  n.trees = 1000, interaction.depth = 2,
                  cv.folds = 10)

# create helper functions
model_type.gbm <- function(x, ...) {
  return("classification")
}

predict_model.gbm <- function(x, newdata, ...) {
  pred <- predict(x, newdata, n.trees = x$n.trees)
  return(as.data.frame(pred))
}

# LIME for two instances in the test dataset

local_obs <- wellbeing_02_test[4:5,] # fourth and fifth instances

explainer_flow <- lime(wellbeing_02[train_flow,], boost_flow)
explanation_flow <- explain(local_obs, explainer_flow, n_labels = 1, n_features = 5)
plot_features(explanation_flow) + ggtitle("LIME")

```

## Train models using caret package

Train a model to predict whether a person falls into high-stress category:

```{r}

# prepare input for caret
wellbeing_01_train_features <- wellbeing_01[train_stress,] %>% select(-HI_STRESS)
wellbeing_01_train_target <- as.data.frame(wellbeing_01[train_stress,])[,"HI_STRESS"]
ntree <- 100

# train
model_rf_stress <- train(x = as.data.frame(wellbeing_01_train_features),
                                y = wellbeing_01_train_target,
                                method = 'rf', ntree = ntree, maximize = FALSE)

```

Train a model to predict hours of flow that a person experience:

```{r}

# prepare input for caret
ntree <- 50
wellbeing_02_train_features <- wellbeing_02[train_flow,] %>%
  select(-FLOW)
wellbeing_02_train_target <- as.data.frame(wellbeing_02[train_flow,])[,"FLOW"]

# train
model_rf <- train(x = as.data.frame(wellbeing_02_train_features),
                      y = wellbeing_02_train_target,
                      method = 'rf', ntree=ntree, maximise = FALSE)

```

## Shapley Values

Shapley values to analyze the prediction of a random forest model predicting whether a person falls into high-stress category.

```{r}

# create prediction from the random forest model generated using caret (see chuck above)
pred_hstress_rf <- Predictor$new(model_rf_stress, data = wellbeing_01_train_features,
                                      class = "1", type = "prob")


# generate shapley values for an indiviual in the dataset 
instance_indices <- 100 # the 100th individual
x.interest <- wellbeing_01_train_features[instance_indices,]

avg_prediction_hstress <-  mean(predict(model_rf_stress, type = 'prob')[,'1'])
actual_prediction_hstress <-  predict(model_rf_stress, newdata = x.interest, type = 'prob')['1']
diff_prediction_hstress  <-  actual_prediction_hstress - avg_prediction_hstress

hstress_shapley_1 <-  Shapley$new(pred_hstress_rf, x.interest = x.interest, sample.size = 100)

# plot
plot(hstress_shapley_1)

```

Shapley values to analyze the prediction of a random forest model predicting hours of flow that a person experience weekly.

```{r}

# create prediction from the random forest model generated using caret (see chuck above)
predictor <- Predictor$new(model_rf, data = as.data.frame(wellbeing_02_train_features))

# generate shapley values for an indiviual in the dataset 
avg_prediction <-  mean(predict(model_rf))
actual_prediction <-  predict(model_rf,
                            newdata = as.data.frame(wellbeing_02_train_features)[instance_indices[2],])
diff_prediction <-  actual_prediction - avg_prediction
x.interest <-  as.data.frame(wellbeing_02_train_features)[instance_indices[2],]

shapley_2 <-  Shapley$new(predictor, x.interest = x.interest)

# plot
plot_shapley_2 <- plot(shapley_2) + 
  scale_y_continuous("Feature value contribution") +
  scale_x_discrete("")

plot_shapley_2 + 
  ggtitle(sprintf("Actual prediction: %.2f\nAverage prediction: %.2f\nDifference: %.2f", 
                  actual_prediction, avg_prediction, diff_prediction))

```

